Pyg4ometry: a Python library for the creation of Monte Carlo radiation transport physical geometries

### 2020-04-10 public/paper/pyg4ometry.tex
Added notes for improvements

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage[T1]{fontenc}
\PassOptionsToPackage{hyphens}{url}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{siunitx}

\usepackage{lineno}
\modulolinenumbers[5]
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

This paper describes such geometry creation and conversion package called {\em pyg4ometry}, implemented in {\em Python} and internally based on the Geant4 geometry 
description. The main features of  {\em pyg4ometry} are a python scripting API to rapidly design parametrised geometry, conversion to and from  FLUKA geometry 
descriptions, conversion from CAD formats  (STEP and IGES) based on FreeCAD \cite{FreeCAD} OpenCascade \cite{OpenCASCADE} and powerful geometry 
visualisation tools based on VTK \cite{VTK4}. The origin of pyg4ometry was a set of utilities to prepare geometry for an accelerator beam line simulation based on Geant4 
called BDSIM. Accelerator physicists like specialists in other areas need a tool to quickly simulate beam losses, and the subsequent interaction of the charged particle 
beam with the accelerator material.

There are existing codes that have functionality similar to pyg4ometry. Tools exist to convert CAD files to GDML, open source examples are GUIMesh \cite{GUIMesh_Pinto}, 
commercial products include ESABASE2 \cite{ESABASE2} and FASTRAD \cite{FASTRAD}. 
CAD software can export data to STL. The existent set of software does not provide a complete set of tools to efficiently create complex geometries.

@@

\subsection{Geant Detector Mark-up Language} 
To exchange geometry descriptions Geant Detector Mark-up Language (GDML) was developed. Geant4 and ROOT can read and write GDML format and it is commonly
used as an exchange format for Geant4 geometries. 
\begin{table}[hbt!]
\centering
\begin{tabular}{|l|l|l|} \hline
GDML tag 		& Child tags				& pyg4ometry class 				\\ \hline
Define 			& Constant				& gdml.Constant				\\
				& Variable					& gdml.Variable					\\
				& Quantity					& gdml.Quantity				\\
				& Expression 				& gdml.Expression				\\
				& Position					& gdml.Position					\\
				& Rotation					& gdml.Rotation				\\
				& Scale					& gdml.Scale					\\
				& Matrix					& gdml.Matrix					\\ \hline
Materials 			& Element					& geant4.Element				\\
				& Isotope					& geant4.Isotope				\\
				& Material					& geant4.Material				\\ \hline
Solid  			& Box					& geant4.solid.Box				\\
	  			& Tube					& geant4.solid.Tubs				\\
	  			& CutTube				& geant4.solid.CutTubs			\\
	  			& Cone					& geant4.solid.Cons				\\
	  			& Para					& geant4.solid.Para				\\
	  			& Trd					& geant4.solid.Trd				\\
	  			& Trap					& geant4.solid.Trap				\\
	  			& Sphere					& geant4.solid.Sphere			\\
	  			& Orb					& geant4.solid.Orb				\\
	  			& Torus					& geant4.solid.Torus				\\
	  			& Polycone				& geant4.solid.Polycone			\\
	  			& GenericPolycone			& geant4.solid.GenericPolycone	\\
	  			& Polyhedra				& geant4.solid.Polyhedra			\\
	  			& Eltube					& geant4.solid.EllipticalTube		\\
	  			& Ellipsoid					& geant4.solid.Ellipsoid			\\
	  			& Elcone					& geant4.solid.EllipticalCone		\\
				& Paraboloid				& geant4.solid.Paraboloid			\\
				& Hype					& geant4.solid.Hyperboloid		\\
				& Tet						& geant4.solid.Tet				\\
				& Xtru					& geant4.solid.ExtrudedSolid		\\
				& TwistedBox				& geant4.solid.TwistedBox		\\
				& TwistedTtap				& geant4.solid.TwistedTrap		\\
				& TwistedTrd				& geant4.solid.TwistedTrd 		\\
				& TwistedTube				& geant4.solid.TwistedTube		\\
				& Arb8					& geant4.solid.GenericTrap		\\
				& Tessellated				& geant4.solid.TessellatedSolid 	\\
				& Union					& geant4.solid.Union				\\
				& Subtraction				& geant4.solid.Subtraction			\\
				& Intersection				& geant4.solid.Intersection 		\\
				& MultiUnion				& geant4.solid.MultiUnion 			\\
				& ScaledSolid				& geant4.solid.ScaledSolid		\\				
	  			& OpticalSurface			& geant4.solid.OpticalSurface		\\ \hline
Structure			& Volume					& geant4.LogicalVolume			\\
				& Assembly				& geant4.AssemblyVolume 		\\
				& PhysVol					& geant4.PhysicalVolume			\\ 
				& ReplicaVol				& geant4.ReplicaVolume			\\ 
				& ParaVol					& geant4.ParameterisedVolume	\\ 
				& DivisionVol				& geant4.DivisionVolume			\\
				& SkinSurface				& geant4.SkinSurface			\\ 
				& BorderSurface			& geant4.BorderSurface			\\ \hline

\end{tabular}
\caption{GDML tags and corresponding pyg4ometry classes.}
\end{table}

\section{pyg4ometry design and layout}
The core of {\em pyg4ometry} consists of python classes that mimic Geant4 solids, logical, physical volume objects, GDML parameters and material classes.
The constructors of the python classes are kept as close to the original Geant4 C++ implementation so that {\it pyg4ometry} users do not have to learn 
a new API. Geometry construction in python proceeds in a way which is very similar to geometry construction in Geant4. A user relatively familiar with Geant4
should be able to start creating geometry in pyg4ometry immediately. In the following sections novel or important developments in {\it pyg4ometry} are described.

@@

\subsection{Disjoint unions}

A disjoint union in FLUKA is a union of two or more zones that are not
connected.  This is a valid construct in FLUKA but is explicitly
forbidden in Geant4 and must be removed during the conversion
process.  Disjoint unions are detected by checking for overlaps
between all zones within a union.

\subsection{Lattice}

Description of how we translate lattices into one LV with multiple PVs.

\begin{itemize}
\item \item disjoint subtractions?
\item quadric specialisation
\item second order lattice
\end{itemize}
\section{Geant4 to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each of the Geant4 solids can be mapped to a FLUKA CSG region.
Some solids in Geant4 directly map to a FLUKA body, others need a construction of a simple FLUKA CSG tree. Table~\ref{tab:geant2fluka}
lists the mapping between geant4 solids and the bodies used to compose a fluka region.

\begin{table}[hbt!]
\centering
\begin{tabular}{| l | l | } \hline
Geant4 solid			& FLUKA region constuction		\\ \hline
Box					& +2 XYP + 2 XZP + 2 YZP 		\\
Tube					& +ZCC - 2 PLA -2 XYP - ZCC	 	\\
CutTube				& +ZCC - 4 PLA -ZCC			\\
Cone				& +TRC - TRC - 2 PLA 			\\
Para					& + 8 PLA						\\
Trd					& + 8 PLA						\\
Trap					& + 8 PLA						\\
Sphere				& +SPH - SPH  - 2 PLA - 2 TRC	\\
Orb					& +SPH						\\
Torus				& +$N$ ZCC  - $N$ PLA			\\
Polycone				& +$N$ TRC -2 PLA				\\
GenericPolycone		& +$N$ ARB					\\
Polyhedra				& +$N$ PLA					\\
Eltube				& +ZEC  - 2 XYP				\\
Ellipsoid				& +ELL - 2 XYP		 			\\
Elcone				& +QUA - 2 XYP				\\
Paraboloid			& +QUA - 2 XYP				\\
Hype					& +QUA - QUA - 2 XYP			\\
Tet					& +4 PLA						\\
Xtru					& +$N_{\rm plane} N_{\rm point}$ PLA \\
TwistedBox			& +$N$ PLA					\\
TwistedTtap			& +$N$ PLA					\\
TwistedTrd			& +$N$ PLA				 	\\
TwistedTube			& +$N$ PLA					\\
Arb8					& +$N$ PLA					\\
Tessellated			& +$N$ PLA				 	\\
Union				& $R_1 \cup R_2$				\\
Subtraction			& $R_1 - R_2$					\\
Intersection			& $R_1 \cap R_2$				\\
MultiUnion			& $R_1 \cup R_2 \cup R_3 \cup$	\\
ScaledSolid			& geant4.solid.ScaledSolid		\\ \hline				
\end{tabular}
\label{tab:geant2fluka}
\caption{GDML/Geant4 solids and the mapping to FLUKA bodies}
\end{table}

@@

\section{Interesting use cases}
pyg4ometry is designed to be as flexible as possible and offer the user a wide range of usage styles and workflows. The following sections describe a selection of 
possible applications.

>>>>

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage[T1]{fontenc}
\PassOptionsToPackage{hyphens}{url}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{siunitx}
\usepackage{lineno}
\modulolinenumbers[5]
\usepackage{xcolor}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

This paper describes such geometry creation and conversion package called {\em pyg4ometry}, implemented in {\em Python} and internally based on the Geant4 geometry 
description. The main features of  {\em pyg4ometry} are a python scripting API to rapidly design parametrised geometry, conversion to and from  FLUKA geometry 
descriptions, conversion from CAD formats  (STEP and IGES) based on FreeCAD \cite{FreeCAD} OpenCascade \cite{OpenCASCADE} and powerful geometry 
visualisation tools based on VTK \cite{VTK4}. The origin of pyg4ometry was a set of utilities to prepare geometry for an accelerator beam line simulation based on Geant4 
called BDSIM \colorbox{yellow}{[CITE]}. Accelerator physicists like specialists in other areas need a tool to quickly simulate beam losses, and the subsequent interaction of the charged particle 
beam with the accelerator material.

There are existing codes that have functionality similar to pyg4ometry. Tools exist to convert CAD files to GDML, open source examples are GUIMesh \cite{GUIMesh_Pinto}, 
commercial products include ESABASE2 \cite{ESABASE2} and FASTRAD \cite{FASTRAD}. There are also tools from the fusion and neutronics community which 
can convert CAD geometry into formats usable by PTMC codes, examples include DagMC \colorbox{yellow}{[CITE]} and McCAD \colorbox{yellow}{[CITE]}.  
CAD software can export data to STL \colorbox{yellow}{[EXPAND]}. The existent set of software does not provide a complete set of tools to efficiently create complex geometries.

@@

\subsection{Geant Detector Mark-up Language} 
To exchange geometry descriptions Geant Detector Mark-up Language (GDML) was developed \colorbox{yellow}{[CITE]}. GDML is an XML based description of 
Geant4 geometry description. Geant4 and ROOT can read and write GDML format and it is commonly used as an exchange format for Geant4 geometries.

\section{Pyg4ometry design and layout}
The core of {\em pyg4ometry} consists of python classes that mimic Geant4 solids, logical, physical volume objects, GDML parameters and material classes.
The constructors of the python classes are kept as close to the original Geant4 C++ implementation so that {\it pyg4ometry} users do not have to learn 
a new API. Geometry construction in python proceeds in a way which is very similar to geometry construction in Geant4. A user relatively familiar with Geant4
should be able to start creating geometry in pyg4ometry immediately. In the following sections novel or important developments in {\it pyg4ometry} are described.

\begin{table}[hbt!]
\centering
\begin{tabular}{ l l l } \hline
GDML tag 		& Child tags				& pyg4ometry class 				\\ \hline
Define 			& Constant				& gdml.Constant				\\
				& Variable					& gdml.Variable					\\
				& Quantity					& gdml.Quantity				\\
				& Expression 				& gdml.Expression				\\
				& Position					& gdml.Position					\\
				& Rotation					& gdml.Rotation				\\
				& Scale					& gdml.Scale					\\
				& Matrix					& gdml.Matrix					\\ \hline
Materials 			& Element					& geant4.Element				\\
				& Isotope					& geant4.Isotope				\\
				& Material					& geant4.Material				\\ \hline
Solid  			& Box					& geant4.solid.Box				\\
	  			& Tube					& geant4.solid.Tubs				\\
	  			& CutTube				& geant4.solid.CutTubs			\\
	  			& Cone					& geant4.solid.Cons				\\
	  			& Para					& geant4.solid.Para				\\
	  			& Trd					& geant4.solid.Trd				\\
	  			& Trap					& geant4.solid.Trap				\\
	  			& Sphere					& geant4.solid.Sphere			\\
	  			& Orb					& geant4.solid.Orb				\\
	  			& Torus					& geant4.solid.Torus				\\
	  			& Polycone				& geant4.solid.Polycone			\\
	  			& GenericPolycone			& geant4.solid.GenericPolycone	\\
	  			& Polyhedra				& geant4.solid.Polyhedra			\\
	  			& Eltube					& geant4.solid.EllipticalTube		\\
	  			& Ellipsoid					& geant4.solid.Ellipsoid			\\
	  			& Elcone					& geant4.solid.EllipticalCone		\\
				& Paraboloid				& geant4.solid.Paraboloid			\\
				& Hype					& geant4.solid.Hyperboloid		\\
				& Tet						& geant4.solid.Tet				\\
				& Xtru					& geant4.solid.ExtrudedSolid		\\
				& TwistedBox				& geant4.solid.TwistedBox		\\
				& TwistedTtap				& geant4.solid.TwistedTrap		\\
				& TwistedTrd				& geant4.solid.TwistedTrd 		\\
				& TwistedTube				& geant4.solid.TwistedTube		\\
				& Arb8					& geant4.solid.GenericTrap		\\
				& Tessellated				& geant4.solid.TessellatedSolid 	\\
				& Union					& geant4.solid.Union				\\
				& Subtraction				& geant4.solid.Subtraction			\\
				& Intersection				& geant4.solid.Intersection 		\\
				& MultiUnion				& geant4.solid.MultiUnion 			\\
				& ScaledSolid				& geant4.solid.ScaledSolid		\\				
	  			& OpticalSurface			& geant4.solid.OpticalSurface		\\ \hline
Structure			& Volume					& geant4.LogicalVolume			\\
				& Assembly				& geant4.AssemblyVolume 		\\
				& PhysVol					& geant4.PhysicalVolume			\\ 
				& ReplicaVol				& geant4.ReplicaVolume			\\ 
				& ParaVol					& geant4.ParameterisedVolume	\\ 
				& DivisionVol				& geant4.DivisionVolume			\\
				& SkinSurface				& geant4.SkinSurface			\\ 
				& BorderSurface			& geant4.BorderSurface			\\ \hline

\end{tabular}
\label{tab:gdml-tags}
\caption{GDML tags and corresponding pyg4ometry classes.}
\end{table}

@@

\subsection{Disjoint unions}
A disjoint union in FLUKA is a union of two or more zones that are not
connected.  This is a valid construct in FLUKA but is explicitly
forbidden in Geant4 and must be removed during the conversion
process.  Disjoint unions are detected by checking for overlaps
between all zones within a union.

\subsection{Lattice}
Description of how we \colorbox{yellow}{translate lattices} 
into one LV with multiple PVs.

\subsection{Example}

\subsection{TODO}
\begin{itemize}
\item disjoint subtractions?
\item quadric specialisation
\item second order lattice
\item Materials 
\end{itemize}
\section{Geant4 to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each of the Geant4 solids can be mapped to a FLUKA CSG region.
Some solids in Geant4 directly map to a FLUKA body, others need a construction of a simple FLUKA CSG tree. Table~\ref{tab:geant2fluka}
lists the mapping between geant4 solids and the bodies used to compose a fluka region.

\begin{table}[hbt!]
\centering
\begin{tabular}{ l  l  } \hline
Geant4 solid			& FLUKA region constuction		\\ \hline
Box					& +2 XYP + 2 XZP + 2 YZP 		\\
Tube					& +ZCC - 2 PLA -2 XYP - ZCC	 	\\
CutTube				& +ZCC - 4 PLA -ZCC			\\
Cone				& +TRC - TRC - 2 PLA 			\\
Para					& + 6 PLA						\\
Trd					& + 6 PLA						\\
Trap					& + 6 PLA						\\
Sphere				& +SPH - SPH  - 2 PLA - 2 TRC	\\
Orb					& +SPH						\\
Torus				& +$N$ ZCC  - $N$ PLA			\\
Polycone				& +$N$ TRC -2 PLA				\\
GenericPolycone		& +$N$ ARB					\\
Polyhedra				& +$N$ PLA					\\
Eltube				& +ZEC  - 2 XYP				\\
Ellipsoid				& +ELL - 2 XYP		 			\\
Elcone				& +QUA - 2 XYP				\\
Paraboloid			& +QUA - 2 XYP				\\
Hype					& +QUA - QUA - 2 XYP			\\
Tet					& +4 PLA						\\
Xtru					& +$N_{\rm plane} N_{\rm point}$ PLA \\
TwistedBox			& +$N$ PLA					\\
TwistedTtap			& +$N$ PLA					\\
TwistedTrd			& +$N$ PLA				 	\\
TwistedTube			& +$N$ PLA					\\
Arb8					& +$N$ PLA					\\
Tessellated			& +$N$ PLA				 	\\
Union				& $R_1 \cup R_2$				\\
Subtraction			& $R_1 - R_2$					\\
Intersection			& $R_1 \cap R_2$				\\
MultiUnion			& $R_1 \cup R_2 \cup R_3 \cup$	\\ \hline
\end{tabular}
\label{tab:geant2fluka}
\caption{GDML/Geant4 solids and the mapping to FLUKA bodies}
\end{table}

@@

\subsection{Non-Convex solid decomposition}
In general the CSG description used by geant4 allow non-convex solids, these include Polyhedron, Xtru, TwistedBox, 
TwistedTrap, TwistedTrd and TwistedTube. These particular solids are problematic when converting geometry to a Fluka 
readable format, as non-convex solids can only be created by union of convex zones. These solids are converted to CGAL 
Nef polyhedra \colorbox{yellow}{XXX CITE} and decomposed to convex polyhedra using a \colorbox{yellow}{XXX} algorithm.

\subsection{Materials}

\subsection{Outstanding issues}
There are still a few outstanding issues with Geant4/GDML to Fluka conversion.

\subsection{Example}

\subsection{TODO}
\begin{itemize}
\item Non-convex solids, convex decomposition (Polyhedron, Xtru, TwistedBox, TwistedTrap, TwistedTrd, TwistedTube)
\item Scaled solid 
\item Placement scale 
\item Replica, Division, Parameterised
\item Materials 
\end{itemize}

\subsection{Example}

\section{Interesting use cases}
Pyg4ometry is designed to be as flexible as possible and offer the user a wide range of usage styles and workflows. The following sections describe a selection of 
possible applications.
